Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-02-27 Ben_Manuscripts/hdphmm/Outline.tex
A few comments.
More implementation details and background

%Predicting Flux and Selectivity of Solutes in a Nanostructured Membrane using the Infinite Hidden Markov Model
\title{Stochastic Models Parameterized by the Infinite Hidden Markov Model Can Predict Flux and Selectivity of Solutes in Nanostructured Membranes.}

\author{Benjamin J. Coscia, Chris Calderon \and Michael R. Shirts}

@@

  We need highly selective membranes in order to perform efficient separations. \\

  \noindent Amphiphilic molecules are capable of self-assembling into ordered nanostructures.\\

  Lyotropic liquid crystals are a class of amphiphilic molecules that can be cross-linked
  into mechanically strong membranes.\\
%  \begin{itemize}
%  	\item H\textsubscript{II} phase lyotropic liquid crystals have densely packed, uniform
%	sized pores and have the potential to disrupt conventional membrane separation
%	techniques by being selective based not only on size and charge, but on chemical
%	functionality as well.
%	\item Q\textsubscript{I} phase LLCs consist of a tortuous network of 3D interconnected
%	pores. They are easier to make.
%  \end{itemize}

  We can only learn so much from experiment. MD can give us mechanistic insights with
  atomistic resolution so that we can intelligently design new membranes for 
  solute-specific separations.\\

@@

  In our previous work, we designed two different approaches which used
  solute time series in order to parameterize stochastic models that could be used
  to project transport on much longer timescales.
  \begin{itemize}
  	\item Brief description of anomalous diffusion
  	\item Brief description of MSDDM
  \end{itemize}

  Although both models had reasonable success at predicting solute MSDs on simulation
  timescales, they had shortcomings.
  \begin{itemize}
  	\item Why MSDDM failed
  	\item Why anomalous diffusion model could be better.
  \end{itemize}

  % BJC: Not sure whether to call it the infinite hidden markov model or hierarchical dirichlet process hidden markov model
  % The former is definitely simpler.
  In this work, we apply the infinite hidden markov model (iHMM), a modeling
  approach that is agnostic to the source of time series data, in order to 
  automatically detect and infer the parameters of an unknown number of latent
  autoregressive modes. \\

  We study the same four solutes from our previous work. \\

  We use the iHMM to generate stochastic trajectory realizations that share the same
  dynamical characteristics as solute trajectories observed in our MD simulations. 
  \begin{itemize}
    \item The trajectories are qualitatively similar, showing expected hopping and trapping
    behavior
    \item They are quantitatively similar in that they reproduce the MSDs measured in MD
  \end{itemize}

  Finally, we use the stochastic trajectory realizations in order to compute the MFPT 
  of solutes in macroscopic length pores.
  \begin{itemize}
  	\item We relate the MFPT to flux and compare the fluxes between solutes to get selectivity.
  \end{itemize}

@@

  \subsection{The Infinite State Hidden Markov Model}\label{method:iHMM}
  %BJC: There is a fair amount of detail here just so I could understand better by writing it out

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the iHMM framework, we estimated the most likely number and sequence of hidden states
  while simultaneously estimating VAR(1) parameters for each state and the overall 
  state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    as shown in SI.
    \item We refer the interested reader to much more extensive descriptions of 
    this process and its implementation. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  We generated stochastic realizations by drawing state sequences based on the
  rows of $T$. 
  %BJC: lots more implementation details once that is ironed out

@@

@@

  \noindent We can relate the identified states back to transport mechanisms.
  \begin{itemize}
  	\item More detailed discussion of identified states
  	\item How size of fluctuations, autoregressive parameters are influenced by trapping mechanisms
  	\item Any new states?
  \end{itemize}

@@

  \noindent We have shown that the iHMM can be used to parameterize solute time series
  with an unknown number of latent dynamical modes. \\

  \noindent We can use the iHMM to help identify mechanisms by relating the latent
  states to observed solute behavior. \\

  \noindent We can use the iHMM to predict macroscopic transport properties. \\

  \noindent The iHMM is not limited to the H\textsubscript{II} phase.

@@

  \newpage

  \section*{TOC Graphic}

>>>>

\title{Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series 
       of Solute Trajectories in Nanostructured Membranes.}

\author{Benjamin J. Coscia, Christopher P. Calderon \and Michael R. Shirts}

@@

  There is a need for highly selective membranes in order to perform efficient 
  separations of components of complex aqueous streams.
  \begin{itemize}
    \item Organic micropollutants
    \item Desalination and boric acid removal from seawater.
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

%  \noindent Amphiphilic molecules are capable of self-assembling into ordered nanostructures.
%  \begin{itemize}
%    \item 
%  \end{itemize}

  Lyotropic liquid crystals (LLC) are a class of amphiphilic molecules whose ordered phases
  can be cross-linked into mechanically strong membranes capable of highly selective
  separations.
  \begin{itemize}
    \item The shape of the LLC monomers and water content dictates the ordered phase 
    that they form. There are two phases of particular interest for membrane applications.
  	\item H\textsubscript{II} phase lyotropic liquid crystals are characterized by 
  	hexagonally packed, straight, pores while the Q\textsubscript{I} phase consists of
  	a tortuous network of 3D interconnected pores. 
  	\item In both cases the pores are uniform in size with radii on the order of 1 nm
  	giving them a very strict molecular weight cut-off.
	\item Additionally, they have the potential to disrupt conventional membrane separation
	techniques by being selective based not only on size and charge, but on chemical
	functionality as well.
	\item Their pores are lined with LLC monomer functional groups which can potentially
	be designed to interact with solutes in a chemically-specific manner
  \end{itemize}

  There are limits to what we can learn from experiment about LLC membrane design.
  \begin{itemize}
    \item Experimental observables like permeability and selectivity allow us to speculate
    about the molecular origins of separation processes.
    \item This drives an empirical design approach which can potentially neglect key
    interactions which influence selectivity.
    \item LLC membranes have been shown to exhibit selectivities which cannot be fully 
    explained by relatively simple macroscopic models. 
  \end{itemize}

  Molecular Dynamics (MD) simulations can give us mechanistic insights with atomistic 
  resolution so that we can intelligently design new membranes for solute-specific 
  separations.
  \begin{itemize}
    \item In our previous work, we built a detailed atomistic model which we used
    to understand the nanoscopic structure of an LLC Membrane.~\cite{coscia_understanding_2019}
    \item We also used the model in order to gain a qualitative understanding of 
    trapping mechanisms which lead to subdiffusive transport behavior.~\cite{coscia_chemically_2019}
  \end{itemize}

@@

  In our previous work, we designed two different approaches which used
  solute time series in order to parameterize stochastic models that could be used
  to project transport on much longer timescales.
  \begin{itemize}
  	\item In our first approach we modeled solute trajectories as subordinated fractional
  	Brownian and L\'evy motion, called the anomalous diffusion (AD) model. 
  	\item We generated solute trajectories by generating a series
  	of anti-correlated hops separated by random periods of entrapment drawn from a 
  	power law distribution.
  	\item Our second approach treated solute motion as a Markov state model with
  	state-dependent dynamics, called the Markov state-dependent dynamical model (MSDDM).
  	\item We parameterized the state transition probabilities between
  	each of eight discrete states as well as the solute dynamics within each of these
  	states. We generated stochastic trajectory realizations by drawing a state
  	sequence based on the transition probability matrix and incorporating the state dynamics
  	while solutes were trapped in each state.
%  	\item Both models had moderate success reproducing the mean squared displacements (MSDs)
%  	exhibited by solutes in our MD simulations.
  \end{itemize}

  Although both models had reasonable success at predicting solute mean squared 
  displacements (MSDs) on MD simulation timescales, they had shortcomings.
  \begin{itemize}
  	\item The MSDDM failed to reproduce the hopping and trapping behavior that
  	characterizes solute center-of-mass trajectories in our MD simulations.
  	\item The AD model did not suffer this qualitative shortcoming, but the 
  	persistent curvature of the predicted MSD curves suggested that the model
  	might underestimate MSDs on long timescales. 
  	\item The formulation of both models required careful examination and
  	characterization of the interactions and dynamics exhibit by MD trajectories 
  	which required considerable human effort.
  \end{itemize}

  % BJC: Not sure whether to call it the infinite hidden markov model or hierarchical dirichlet process hidden markov model
  % The former is definitely simpler.
  % MRS: True.  I would probably see which is used more in the literature. 
  % BJC1: going with IHMM. HDPHMM seems to mostly be used by Emily Fox
  In this work, we apply the infinite hidden markov model (IHMM), a modeling
  approach that is agnostic to the source of time series data, in order to 
  automatically detect and infer the parameters of an unknown number of latent
  autoregressive (AR) modes present in solute center-of-mass time series.
  \begin{itemize}
  	\item In addition to AR parameters for each state, the IHMM estimates the
  	state transition probability matrix.
  	\item The model helps simultaneously uncover underlying transport mechanisms which
  	give rise to dynamical behavior and project that behavior on longer timescales so
  	that we can estimate macroscopic transport observables.
  \end{itemize}

  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms.
  \begin{itemize}
    \item We compare the inferred mechanisms to those which we manually
    identified in our previous work. 
    \item Some kind of conclusion here. Did we find more or less states. Any new states/ subdivisions of states?
  \end{itemize}

  We can also use the IHMM to generate stochastic trajectory realizations that share
  the same dynamical characteristics as solute trajectories observed in our MD 
  simulations. 
  \begin{itemize}
    \item The trajectories are qualitatively similar, showing expected hopping and trapping
    behavior.
    \item They are quantitatively similar in that they reproduce the MSDs measured in MD.
  \end{itemize}

  Finally, we use the stochastic trajectory realizations in order to compute the 
  macroscopic flux of each solute and selectivity of the LLC membranes studied towards
  each solute.
  \begin{itemize}
    \item We relate these macroscopic properties to our nanoscopic model by simulating 
    mean first passage time (MFPT).
    \item Some kind of conclusion. This membrane is selective towards solutes with this
    functionality. 
    \item Does the conclusion agree with our previous work? Any length dependence? (I think not)
  \end{itemize}

@@

  \subsection{The Infinite State Hidden Markov Model}\label{method:IHMM}
  %BJC: There is a fair amount of detail here just so I could understand better by writing it out

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden states
  while simultaneously estimating VAR(1) parameters for each state and the overall 
  state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    as shown in SI.
    \item We refer the interested reader to much more extensive descriptions of 
    this process and its implementation. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  We applied the IHMM algorithm to the radial and axial coordinates of each 
  solute center-of-mass trajectory.
  \begin{itemize}
    \item We measured the radial coordinate as the distance from the closest
    pore center.
    \item Because the pores are somewhat tortuous, we approximated the pore 
    center as a spline. See previous work.
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
    \item Since the mean vector, $\mathbf{c}$ of each state can take on 
    continuous values, the IHMM algorithm tends to find a large number of states,
    with very few states revisited.
    % BJC1: This will need to be updated depending on how I cluster. Might need to describe
    % multiple clustering approaches.
    \item However, $\mathbf{A_i}$ and $\mathbf{\Sigma}$ tend to share similarities with
    independently identified states. Therefore, we clustered based on $\mathbf{A_i}$
    and $\mathbf{\Sigma}$, ignoring the mean.
    \item Since  we do not know the number of similar states beforehand, we used
    a non-parametric Bayesian Gaussian mixture model % cite scikit-learn
    \item We used the clustered parameters in order to relabel states in the estimated
    state sequences and to recalculate the state transition probability matrix.
  \end{itemize}

  We generated stochastic trajectory realizations by drawing state sequences 
  based on the rows of $T$.
  \begin{itemize}
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item We set the unconditional mean of each state based on the position 
    before the state transition occurred.
  \end{itemize}

@@

  \subsection{Inferring Solute Transport Mechanisms}

  Clustering parameters sets results in X distinct dynamical modes.
  \begin{itemize}
    \item In the figure below, we show time series simulations that qualitatively
    illustrate the difference in dynamical behavior between modes.
  \end{itemize}

  %BJC1: a figure showing some representative fluctuations in each mode

  \noindent We can relate the identified states back to transport mechanisms.
  \begin{itemize}
  	\item More detailed discussion of identified states
  	\item How size of fluctuations, autoregressive parameters are influenced by trapping mechanisms
  	\item How do these states compare to those identified in our previous work?
  	\item Any new states?
  \end{itemize}

  \subsection{Reproducing MD Trajectories and MSDs with the IHMM}

@@

  \subsection{Estimating Solute Flux and Selectivity}

@@

  \noindent We have shown that the IHMM can be used to parameterize solute time series
  with an unknown number of latent dynamical modes. \\

  \noindent We can use the IHMM to help identify mechanisms by relating the latent
  states to observed solute behavior. \\

  \noindent We can use the IHMM to predict macroscopic transport properties. \\

  \noindent The IHMM is not limited to the H\textsubscript{II} phase.

@@

  %\newpage

  %\section*{TOC Graphic}
